Customizing Software by Direct Manipulation of Tabular Data

### 2020-05-22 paper.tex
full draft

Most attempts at empowering end users to customize their software offer
a simplified version of programming. Some scripting languages
\citep{bolin2005, cook2007} have a friendly syntax that resembles
natural language. Some visual customization tools eliminate text syntax
entirely. Macro recorders \citep{cook2007, chasins2018, anupam2000}
remove some of the initial programming burden by letting a user start
with concrete demonstrations.

Despite their many differences, these approaches all share something in
common: an imperative programming model, with statement sequencing,
mutable variables and loops. End users express their ideas in
scripts---sequences of commands---which, name aside, are not very
different from conventional code.

@@

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:examples}, we introduce the key
ideas of table-driven customization through an example scenario.

@@

\hypertarget{sec:examples}{%
\section{Example Scenario}\label{sec:examples}}

To concretely illustrate the end user experience of table-driven
customization, here is a real example of using the Wildcard browser
extension to customize a website.

\begin{itemize}
\tightlist
\item
  Hacker News
\item
  open the table
\item
  sort by comments
\item
  take notes on links
\item
  filter out visited sites with a formula
\item
  add read times with a formula
\end{itemize}

@@

@@

Figure~\ref{fig:table-adapter} summarizes the overall architecture of
table-driven customization, using a simplified version of the Airbnb
example above. In this example, the name of each listing is scraped from
the web page DOM, the latitude and longitude of each listing is scraped
from AJAX responses, and user annotations are loaded from the brower's
local storage.

@@

\textbf{Information from other tables}: The query engine sends each
table adapter the entire combined table of data being rendered to the
user. The DOM scraping adapter uses this to add annotations to the page
by looking for additional data columns joined to scraped rows, and
rendering them in the page.

\textbf{Currently selected record}: The query engine notifies each table
adapter about which record is currently selected by the user in the
table UI. The DOM scraping adapter uses this information to highlight
the row in the page that corresponds to the selected record in the
table.

@@

Here we describe in more detail the table adapters that we have
implemented in Wildcard to power the customizations shown in
Section~\ref{sec:examples}.

@@

We have also found instances where simple data manipulation is useful:
adding together different subtotals into a total price, or manipulating
the results of a formula expression with arithmetic or concatenating
string labels, as shown in the example in Section~\ref{sec:examples}.
(\emph{todo: this makes sense iff we switch the examples sectino to HN})

@@

``Cell editors'' are UI widgets that expose a custom editing UI for a
single cell of the table view. A programmer building a cell editor need
only integrate it with the table viewer; propagating values into the
website UI is handled by the site-specific DOM adapter. As a result,
creating a new cell editor can be as easy as just writing some glue code
between the table editor and an existing UI widget library.

@@

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for a gentler slope in customization, and we show
in Section~\ref{sec:examples} and Section~\ref{sec:evaluation} that our
model can still implement many useful customizations in practice. Also,
there is sometimes a way to reframe an imperative script in terms of our
direct manipulation model; for example, a script that iterates through
rows in a page adding some additional information can be reproduced
using a formula in Wildcard.

@@

Software customization tools typically fall into one of two categories:
third-party, or semantic.

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\emph{todo}

>>>>

Many applications don't meet the precise needs of their users. End user
customization systems can help improve the situation, by empowering
non-programmers to modify their software in the way they would like.

Most end user customization systems offer a simplified version of
programming. Some scripting languages \citep{bolin2005, cook2007} have a
friendly syntax that resembles natural language. Some visual
customization tools eliminate text syntax entirely. Macro recorders
\citep{cook2007, chasins2018, anupam2000} remove some of the initial
programming burden by letting a user start with concrete demonstrations.

Despite their many differences, these approaches all share something in
common: an imperative programming model, with statement sequencing,
mutable variables and loops. End users express their ideas in scripts
made up of sequences of commands.

@@

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:example}, we introduce the key
ideas of table-driven customization through an example scenario.

@@

\hypertarget{sec:example}{%
\section{Example Scenario}\label{sec:example}}

To illustrate the end user experience of table-driven customization, we
consider a real example of using the Wildcard browser extension to
customize Hacker News, a tech news aggregator. Figure 1 shows
accompanying screenshots.

@@

When the user opens Hacker News in a browser equipped with the Wildcard
extension, they see a table at the bottom of the page, listing
information about each link on the site: its title, URL, number of
upvote points, etc. (Figure~\ref{fig:hacker-news}, Note A) The end user
didn't need to do any work to produce this table; Wildcard contains a
library of adapters created by programmers for extracting data from
websites.

The user's first goal might be to change the ranking of links on the
homepage. Hacker News itself uses a ranking algorithm in which the
position of an article depends not only on its point count (a measure of
popularity), but also on how long it has been on the site. If the user
hasn't been checking the site frequently, it'seasy to miss a popular
article that has fallen down the page. Sorting the page solely by number
of points would achieve a more stable ranking, with the most upvoted
articles remaining at the top of the list until they are no longer on
the home page.

To achieve this sort order, the user simply clicks on the ``points''
column header in the table. This sorts the table view by points, and the
website UI also becomes sorted in the same order
(Figure~\ref{fig:hacker-news}, Note B). This sort order persists across
page loads, so every time the user loads Hacker News, they will see
articles with this stable ordering.

Next, the user might decide to attempt a more substantial customization:
to show estimated read times next to each article, as additional context
when deciding what to read. The table contains additional columns to the
right of the original attributes for each link, where the user can enter
spreadsheet-style formulas to compute derived values. The user enters a
formula (Figure~\ref{fig:hacker-news}, Note C):

\begin{verbatim}
=ReadTimeInSeconds(link)
\end{verbatim}

This calls a built-in function that takes a URL as an argument and makes
a request to a public web service that returns an estimated read time
for the contents of that link. The result of evaluating the formula is
simply a numerical value in seconds, which is not a particularly legible
format. The user enters another formula in the next column
(Figure~\ref{fig:hacker-news}, Note D):

\begin{verbatim}
=Concat(Round(user1/60), "min read")
\end{verbatim}

This converts seconds to minutes by dividing by 60 and rounding to the
nearest integer, and then concatenates a label to the number. Finally,
the user chooses to show the results of this new column in the original
page (Figure~\ref{fig:hacker-news}, Note E). Each article on the page
now shows an annotation with the estimated read time in minutes.

The user might also manually add notes to the table without using
formulas. In this case, the user jots down a few notes about articles
they might want to read, and the notes appear in the page as well.

The features presented in this scenario can support a broad range of
other customizations. Section~\ref{sec:evaluation} explains many other
use cases, but first we explain how the system works internally.

@@

Figure~\ref{fig:table-adapter} summarizes the overall architecture of
table-driven customization, using a simplified version of the Airbnb
example above. (\emph{todo: change this figure to hacker news}) In this
example, the name of each listing is scraped from the web page DOM, the
latitude and longitude of each listing is scraped from AJAX responses,
and user annotations are loaded from the brower's local storage.

@@

\textbf{Information from other tables}: The query engine sends each
table adapter the entire combined table of data being rendered to the
user. The DOM scraping adapter uses this to add annotations to the page
by looking for additional data columns joined to scraped rows, and
rendering them in the page. (\emph{todo: explain joined})

\textbf{Currently selected record}: The query engine notifies each table
adapter about which record is currently selected by the user in the
table UI. The DOM scraping adapter uses this information to highlight
the row in the page that corresponds to the selected record in the
table. (\emph{todo: clarify})

@@

Here we describe in more detail the table adapters that we have
implemented in Wildcard to power the customizations shown in
Section~\ref{sec:example}.

@@

We have also found instances where simple data manipulation is useful:
adding together different subtotals into a total price, or manipulating
the results of a formula expression with arithmetic or concatenating
string labels, as shown in the example in Section~\ref{sec:example}.
(\emph{todo: this makes sense iff we switch the examples sectino to HN})

@@

Cell editors are UI widgets that expose a custom editing UI for a single
cell of the table view. A programmer building a cell editor need only
integrate it with the table viewer; propagating values into the website
UI is handled by the site-specific DOM adapter. As a result, creating a
new cell editor can be as easy as just writing some glue code between
the table editor and an existing UI widget library.

@@

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for a gentler slope in customization, and we show
in Section~\ref{sec:example} and Section~\ref{sec:evaluation} that our
model can still implement many useful customizations in practice. Also,
there is sometimes a way to reframe an imperative script in terms of our
direct manipulation model; for example, a script that iterates through
rows in a page adding some additional information can be reproduced
using a formula in Wildcard.

@@

Software customization tools typically fall into one of two categories:
ad hoc, or semantic.

@@

\hypertarget{sec:conclusion}{%
\section{Conclusion and Future Work}\label{sec:conclusion}}

In this paper, we have presented table-driven customization, a new
paradigm for customizing software by direct manipulation of the
underlying structured data. We have demonstrated the paradigm using a
browser extension, and have used it to create useful customizations for
a variety of websites.

Our primary goal for future work is to evaluate the system with a
broader group of users. What barriers do end users face when using the
system? What types of customizations do they choose to create? What
formulas prove most helpful?

Another area for future work is exploring whether it is possible to
express a broader range of customizations by extending our table-editing
paradigm. Are there ways to offer an increase in power and functional
complexity, while retaining a programming model that is simpler for end
users than conventional coding?

As computing plays an ever greater role in our lives, it is increasingly
important that end users have agency over the behavior of their
software, rather than having every detail be dictated by companies whose
incentives are not always aligned with the user's interests. We hope
that table-driven customization can serve as one point, and maybe a
guidepost, on our path from normal use to deep modification, in support
of a more adaptable experience for all computer users.
